---
title: "Home Team Advantage in International Soccer"
author: "Michael Nichols, Evan Weller"
date: "2024-12-11"
output: 
  pdf_document:
    toc: true
    number_sections: true
---

## Introduction:

### Research Question

Does a 'home field advantage' exist in international soccer?

### Data Source

The data, *International Football Results from 1872 to 2024*, was sourced from [Kaggle](https://www.kaggle.co/datasets/martj42/international-football-results-from-1872-2017). This data set contains the results and statistics from over 40,000 international soccer matches from as far back as 1872.

#### Provenance

The dataset was collected by Mart Jürisoo to analyze international football trends and performance. It includes matches from over 200 nations, covering friendlies, qualifiers, and tournament matches.

#### FAIR Principles

The dataset adheres to FAIR principles: - **Findable:** Available on Kaggle with descriptive metadata. - **Accessible:** Publicly downloadable in CSV format. - **Interoperable:** Compatible with standard data analysis tools. - **Reusable:** Clear licensing allows academic and non-commercial usage.

#### CARE Principles

The dataset does not involve sensitive or personal data, making ethical concerns minimal. It focuses on publicly available sports data.

## Results and Discussion

### Home Team Win Percentage

Figure 1 illustrates the win percentage of home teams across all matches. The home team wins approximately **X%** of the time.

### Average Goals Scored

Figure 2 compares average goals scored by home teams (**Y goals**) against the average total goals in matches (**Z goals**). This suggests...

------------------------------------------------------------------------

## Data Visualizations

```{r include = FALSE}
#|label = Cleaning Data
# Load libraries

library(tidyverse)
library(knitr)

# Load data
football_data <-read_csv("results.csv")

# Filter out neutral matches
football_clean <- football_data %>%
  filter(neutral == FALSE)

# Add match outcome column
football_clean <- football_clean %>%
  mutate(
    outcome = case_when(
      home_score > away_score ~ "Home Win",
      home_score < away_score ~ "Away Win",
      TRUE ~ "Draw"
    )
  )

# Home win percentage
home_win_rate <- as.integer(mean(football_clean$outcome == "Home Win") * 100)

# Average goals
avg_home_goals <- mean(football_clean$home_score)
avg_total_goals_per_team <- mean((football_clean$home_score + football_clean$away_score)/2)
```

```{r echo = FALSE}
football_clean %>%
  select(home_team, away_team, home_score, away_score, outcome, date) %>%
  head(10) %>%  # Display the first 10 rows for clarity
  kable(caption = "Football Match Results (Filtered)")
```

### Figure 1: Home Team Win Percentage

A bar plot of match outcomes (win/loss/draw) categorized by match type.

```{r include = FALSE}
#|Label = Figure 1
# Plot 1: Home team win percentage
library(ggplot2)

win_plot <- football_clean %>%
  count(outcome) %>%
  ggplot(aes(x = outcome, y = n, fill = outcome)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Match Outcome",
    x = "Match Outcome",
    y = "Count"
  ) +
  theme_minimal()
ggsave("win_plot.pdf", win_plot, width = 8, height = 6)
```

```{r echo = FALSE}
print(win_plot)
```

### Figure 2: Average Goals Comparison

A side-by-side bar chart comparing home goals and total match goals.

```{r include = FALSE}
#|Label = Figure 2
# Plot 2: Average goals comparison
goals_plot <- tibble(
  Type = c("Home Goals", "Total Goals Per Team"),
  Average = c(avg_home_goals, avg_total_goals_per_team)
) %>%
  ggplot(aes(x = Type, y = Average, fill = Type)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Average Goals Scored",
    x = "Home Goal or Team Goal",
    y = "Goals"
  ) +
  theme_minimal()
ggsave("goals_plot.pdf", goals_plot, width = 8, height = 6)
```

```{r echo= FALSE}
print(goals_plot)
```

------------------------------------------------------------------------

## Conclusion

------------------------------------------------------------------------

## Appendix: Code

```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```
